Topic categorization in Portuguese news articles

### 2018-01-26 paper/article.tex
camera ready

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{textcomp}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{siunitx}
\usepackage[T1]{fontenc}
\usepackage[utf8x]{inputenc}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\lstset{basicstyle=\footnotesize\ttfamily,breaklines=true,literate={á}{{\'a}}1 {ê}{{\^e}}1 {é}{{\'e}}1 }

\title{Topic categorization in Portuguese news articles\\
\thanks{Fundação para a Ciência e Tecnologia} }

\author{\IEEEauthorblockN{André F. Santos}
\IEEEauthorblockA{\textit{CRACS \& INESC-Porto LA} \\ \textit{Faculty
of Sciences, University of Porto}\\ Porto, Portugal \\
afs@inesctec.pt} }

@@

\begin{IEEEkeywords} topic categorization, machine learning, text
mining \end{IEEEkeywords}

\section{Introduction}
Online news articles first appeared as reprints from traditional
newspapers; nowadays, however, they represent now the primary source
of news for some segments of the population, both in developed and
developing countries (whether consumed directly in the newspaper
website, or indirectly e.g. through a social media application or a
feed catcher)\cite{greer2006evolution,boczkowski2005digitizing,chyi1999access}.

Unofficially know as \emph{the fourth branch of government}, the press
plays a vital role within our society, keeping us informed about
the current state of affairs (at a local and global scale) and acting
as a watchdog for the other three branches (legislative, executive and
judicial). The (lack of) freedom of press and access to the news in a
given country is even often considered an indicator of a lack of
democracy\cite{goode2009social,house2009freedom}.

As such, improving the ways citizens can access the information (view
it, query it and search it) contained in news articles has the
potential to contribute for a more informed and, ultimately, a
better society\cite{bollinger1988tolerant}.

On the other hand, the last decades have witnessed a fast increase on
the rate of publication of digital text documents. Traditional
document types, such as news articles, scientific papers or books are
now published online together with new formats, such as blog posts or
tweets, each having thousands or millions of new documents published
each day\cite{hilbert2011world,allan2006online}.

And it was not only the publication step which has moved to the
digital world; in fact, most often nowadays the whole document
lifecycle happens digitally, with virtual tools available for
researching, writing, styling, publishing and sharing\cite{williams2009personal}.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=.97\columnwidth]{imgs/wireframe}}
\caption{Category classification and suggestion on the client side}
\label{fig:wireframe}
\end{figure}

The simplest way of achieving this categorization is to have the
author of the article to manually introduce it (e.g. the journalist
typing it on the news article authoring framework); however, this
solution presents some challenges:
\begin{itemize}
    \item it increases the amount of work the author has to do
    \item the author might not be sure which categories are available
    \item the author might not be sure which category is the best
        (e.g. \emph{Economics} vs  \emph{Finance})
    \item it does not scale -- e.g. if the goal is to categorize an
        existing (large) corpus
\end{itemize}

@@

\section{Methods}
In order to train and evaluate classification algorithms, we first
needed to choose and obtain a suitable dataset. Preferably, this
dataset would contain documents were already categorized, which would
allows to skip the time and effort-consuming step of manually
categorizing the articles ourselves. Once this dataset was chosen and
obtained, we would then clean and prepare it to be used to train the
classifiers.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=.97\columnwidth]{imgs/articles}}
\caption{Total number of articles retrieved for each category}
\label{fig:articles}
\end{figure}

@@

Originally, the dataset was obtained as large MongoDB collection (more
than 2.5 million entries), containing articles from several Portuguese
and international newspapers. The process needed to transform this
collection into data our classifiers could process required querying
the database, exporting the news articles and splitting them into a
train and an evaluation datasets.

@@

The files were then stored in folders, separated by the category.
Furthermore, for each category, 700 articles were allocated to the
train set, and 200 to the evaluation set. These preprocessing tasks
were accomplished using Bash and Node.js scripts.

Both datasets were then loaded into R using the
\texttt{tm}\footnote{\url{https://cran.r-project.org/web/packages/tm/}}
package. Each was then passed to a function responsible for
preprocessing the text of the articles:

\begin{itemize}
    \item the text was converted to lowercase characters
    \item the Portuguese stopwords were removed
    \item diacritics were converted to their normalized form (e.g.
        \framebox{à} $\rightarrow$ \framebox{a})
    \item punctuation signs were removed
    \item numbers were removed
    \item word were stemmed (e.g. \framebox{conseguiram} $\rightarrow$
        \framebox{consegu})
    \item whitespace was removed and the text was tokenized
\end{itemize}

A document-term matrix was then calculated using the
\texttt{DocumentTermMatrix} method from the \texttt{tm} package.
For each of the 5600 documents preset in the train set the matrix
included all the terms which met the following requirements:

\begin{itemize}
    \item the term length was between 3 and 30 characters (to discard
        things like URLs or badly tokenized sentences)
    \item the term appeared in the document at least twice
    \item the term appeared at least in 10 documents
\end{itemize}
The algorithm used to weigh the terms in the matrix was the
\textit{tf-idf}\cite{christopher2008introduction}. The obtained matrix
presented a sparsity of 99\% and contained 3234 distinct terms.

@@

We used the \texttt{removeSparseTerms} function from the \texttt{tm}
package, and produced three distinct lists of terms:
\begin{description}
    \item[$LT_{90}$] contained terms with 90\% or less sparsity
    \item[$LT_{95}$] contained terms whose sparsity was under 95\%
    \item[$LT_{99}$] contained the terms with a sparsity level below 99\%
\end{description}

@@

To create the classifiers, we used 5 well knows algorithms: decision
tree (DT), k-nearest neighbors (KNN), naive Bayes (NB), neural network
(NN) and support vector
machine -- with radial kernel (SVM\textsubscript{RK}) and linear
kernel (SVM\textsubscript{LK}).

For the process of training a

@@

\begin{table}[htbp]
    \vspace{-.5cm}
    \caption{Lists size}
\begin{center}
\begin{tabular}{l|SS}
    & \parbox{6em}{\centering Sparsity\\threshold (\%)}  &
    \parbox{5em}{\centering Number\\of terms}   \\\hline
    $LT_{90}$ &  90                 &  45 \\
    $LT_{95}$ &  95                 & 139 \\
    $LT_{99}$ &  99                 & 912 \\
\end{tabular}
\label{tab:size}
\end{center}
\end{table}

@@

\begin{table}[htbp]
    \vspace{-.2cm}
    \caption{Execution times for training}
\begin{center}
\begin{tabular}{l|cr}
    & Algorithm  & \parbox{3em}{\centering Training\\time} \\\hline
    $LT_{90}$    & DT         & 12s  \\
    $LT_{95}$    & KNN        &  3m  \\
    $LT_{99}$    & KNN        & 57m  \\
\end{tabular}
\label{tab:exectime}
\end{center}
\end{table}

@@

\begin{table}[htbp]
    \vspace{-.2cm}
    \caption{Error rates}
\begin{center}
\begin{tabular}{l|rrrrrr}
                 & DT            & KNN           & NB   & NN   & SVM\textsubscript{RK} & SVM\textsubscript{LK} \\ \hline
    $LT_{90}$    & \textbf{0.60} & 0.71          & 0.85 & 0.61 & 0.88                  & 0.88                  \\
    $LT_{95}$    & 0.70          & \textbf{0.56} & 0.83 & 0.58 & 0.87                  & 0.87                  \\
    $LT_{99}$    & 0.70          & \textbf{0.35} & 0.87 & 0.76 & 0.87                  & 0.87                  \\
\end{tabular}
\label{tab:error}
\end{center}
\end{table}

@@

\begin{table}[htbp]
    \caption{Categories confusion matrix ($LT_{99}$ with KNN)}
\begin{center}
\begin{tabular}{l|rrrrrrrr}
& \rotatebox{75}{science}    & \rotatebox{75}{culture}  &
    \rotatebox{75}{sports} & \rotatebox{75}{economics} &
    \rotatebox{75}{international}    & \rotatebox{75}{music}   &
    \rotatebox{75}{national}     & \rotatebox{75}{politics} \\\hline
science       & {\bf 127} & 19        & 1          & 6         & 12       & 24        & 11       & 0 \\
culture       & 5         & {\bf 102} & 2          & 2         & 3        & 79        & 6        & 1 \\
sports        & 1         & 3         & {\bf  168} & 7         & 3        & 14        & 4        & 0 \\
economics     & 5         & 3         & 0          & {\bf 146} & 5        & 17        & 14       & 10 \\
international & 12        & 12        & 7          & 17        & {\bf 90} & 35        & 16       & 11 \\
music         & 0         & 2         & 0          & 0         & 2        & {\bf 184} & 1        & 0 \\
national      & 9         & 9         & 9          & 19        & 15       & 31        & {\bf 88} & 21 \\
politics      & 1         & 3         & 0          & 30        & 8        & 17        & 14       & {\bf 127} \\
\end{tabular}
\label{tab:confmat}
\end{center}
\end{table}

@@

The size of the list of terms, for example, gives us an idea of the
memory footprint of a classifier, a parameter which is of the utmost
importance if the goal is to implement a classifier as lightweight as
possible. The training time is also relevant, as shorter training
times give the possibility of retraining the classifiers more often,
allowing them to be updated as the corpus of news articles grows.

Looking at the actual results obtained and represented in
Tables~\ref{tab:size} and \ref{tab:exectime} we can see that $LT_{90}$
has simultaneously the smallest list of terms (45) and the shortest
training time (12 seconds using the decision tree algorithm). However,
$LT_{90}$ also presents the worst error rates, even when looking at
the algorithm which achieved its best results (0.60 using a decision
tree classifier).

@@

The results obtained confirm that there is a tradeoff between the size
of the list of terms and the training time, on the one hand, and the
classifier error rate, on the other. However, a list of 912 terms
seems to be an acceptable memory footprint for a client side
classifier; additionally, the additional training time would not present
much problems in this scenario, as the classifier would be trained
before hand and thus would not be visible on the client side.

Given the reasonable values for the training time and the size of the
list of terms, and looking to the error rate obtained in each of tests
performed, we can conclude that the best option was to use the largest
list of terms ($LT_{99}$) and the k-nearest neighbors classifier.

@@

The tasks and results previously described already provide useful
insights into this matter. However, the lowest error rate obtained
(0.35), despite already acceptable, might still be improved, either by
trying new algorithms, fine tuning the ones already tested, or by
increasing the train corpus.

We have established that it is in fact possible
to develop a news article classifier which is lightweight enough to be
used (in real time) on the client side of a web application. The
following step will be the actual implementation, probably in the form
of a JavaScript library, of such a classifier.

@@

\bibliography{article}
\bibliographystyle{splncs03}

>>>>

% This template has been tested with LLNCS DOCUMENT CLASS -- version 2.20 (24-JUN-2015)

% !TeX spellcheck = en-US
% !TeX encoding = utf8
% !TeX program = pdflatex
% !BIB program = bibtex
% -*- coding:utf-8 mod:LaTeX -*-

%"runningheads" enables:
%  - page number on page 2 onwards
%  - title/authors on even/odd pages
%This is good for other readers to enable proper archiving among other papers and pointing to
%content. Even if the title page states the title, when printed and stored in a folder, when
%blindly opening the folder, one could hit not the title page, but an arbitrary page. Therefore,
%it is good to have title printed on the pages, too.
\documentclass[runningheads,a4paper]{llncs}[2015/06/24]

%cmap has to be loaded before any font package (such as cfr-lm)
\usepackage{cmap}

\iftrue % use default-font
  %better font, similar to the default springer font
  %cfr-lm is preferred over lmodern. Reasoning at http://tex.stackexchange.com/a/247543/9075
  \usepackage[%
    rm={oldstyle=false,proportional=true},%
    sf={oldstyle=false,proportional=true},%
    tt={oldstyle=false,proportional=true,variable=true},%
    qt=false%
  ]{cfr-lm}
\else
  %
  %if more space is needed, exchange cfr-lm by two packages from newtx:
  \usepackage{newtxtext}
  \usepackage{newtxmath}
  \usepackage[zerostyle=b,scaled=.9]{newtxtt}
\fi

%for demonstration purposes only
\usepackage[math]{blindtext}

%% If you need packages for other papers,
%% START COPYING HERE
%%
%% COPY ALSO cmap from line 19 - it has to be loaded as the first package

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc} %support umlauts in the input

\usepackage{graphicx}

%Set English as language and allow to write hyphenated"=words
%
%Even though `american`, `english` and `USenglish` are synonyms for babel package (according to https://tex.stackexchange.com/questions/12775/babel-english-american-usenglish), the llncs document class is prepared to avoid the overriding of certain names (such as "Abstract." -> "Abstract" or "Fig." -> "Figure") when using `english`, but not when using the other 2.
%english has to go last to set it as default language
\usepackage[ngerman,english]{babel}
%Hint by http://tex.stackexchange.com/a/321066/9075 -> enable "= as dashes
\addto\extrasenglish{\languageshorthands{ngerman}\useshorthands{"}}

%Sorts the citations in the brackets
%It also allows \cite{refa, refb}. Otherwise, the document does not compile.
%  Error message: "White space in argument"
\usepackage{cite}

% backticks (`) are rendered as such in verbatim environment. See https://tex.stackexchange.com/a/341057/9075 for details.
\usepackage{upquote}

%extended enumerate, such as \begin{compactenum}
\usepackage{paralist}

%put figures inside a text
%\usepackage{picins}
%use
%\piccaptioninside
%\piccaption{...}
%\parpic[r]{\includegraphics ...}
%Text...

%for easy quotations: \enquote{text}
\usepackage{csquotes}

%enable margin kerning
\RequirePackage{iftex}
\ifPDFTeX
  \RequirePackage[%
    final,%
    tracking=smallcaps,%
    expansion=alltext,%
    protrusion=alltext-nott]{microtype}%
\else
  \RequirePackage[%
    final,%
    protrusion=alltext-nott]{microtype}%
\fi%

%tweak \url{...}
\usepackage{url}
%\urlstyle{same}
%improve wrapping of URLs - hint by http://tex.stackexchange.com/a/10419/9075
\makeatletter
\g@addto@macro{\UrlBreaks}{\UrlOrds}
\makeatother
%nicer // - solution by http://tex.stackexchange.com/a/98470/9075
%DO NOT ACTIVATE -> prevents line breaks
%\makeatletter
%\def\Url@twoslashes{\mathchar`\/\@ifnextchar/{\kern-.2em}{}}
%\g@addto@macro\UrlSpecials{\do\/{\Url@twoslashes}}
%\makeatother

%diagonal lines in a table - http://tex.stackexchange.com/questions/17745/diagonal-lines-in-table-cell
%slashbox is not available in texlive (due to licensing) and also gives bad results. This, we use diagbox
%\usepackage{diagbox}

%required for pdfcomment later
\usepackage{xcolor}

%for listings
\usepackage{listings}
\lstset{%
  basicstyle=\ttfamily,%
  columns=fixed,%
  basewidth=.5em,%
  xleftmargin=0.5cm,%
  literate={á}{{\'a}}1 {ê}{{\^e}}1 {é}{{\'e}}1,
  captionpos=b}%
\renewcommand{\lstlistingname}{List.}
%Fix counter as described at https://tex.stackexchange.com/a/28334/9075
\usepackage{chngcntr}
\AtBeginDocument{\counterwithout{lstlisting}{section}}

%compatibility of packages minted and listings with respect to the numbering of "List." caption
%source: https://tex.stackexchange.com/a/269510/9075
\AtBeginEnvironment{listing}{\setcounter{listing}{\value{lstlisting}}}
\AtEndEnvironment{listing}{\stepcounter{lstlisting}}

%enable nice comments
%this also loads hyperref
\usepackage{pdfcomment}
%enable hyperref without colors and without bookmarks
\hypersetup{hidelinks,
  colorlinks=true,
  allcolors=black,
  pdfstartview=Fit,
  breaklinks=true}
%enables correct jumping to figures when referencing
\usepackage[all]{hypcap}

\newcommand{\commentontext}[2]{\colorbox{yellow!60}{#1}\pdfcomment[color={0.234 0.867 0.211},hoffset=-6pt,voffset=10pt,opacity=0.5]{#2}}
\newcommand{\commentatside}[1]{\pdfcomment[color={0.045 0.278 0.643},icon=Note]{#1}}

%compatibality with packages todo, easy-todo, todonotes
\newcommand{\todo}[1]{\commentatside{#1}}
%compatiblity with package fixmetodonotes
\newcommand{\TODO}[1]{\commentatside{#1}}

%enable \cref{...} and \Cref{...} instead of \ref: Type of reference included in the link
\usepackage[capitalise,nameinlink]{cleveref}
%Nice formats for \cref
\crefname{section}{Sect.}{Sect.}
\Crefname{section}{Section}{Sections}
\crefname{listing}{\lstlistingname}{\lstlistingname}
\Crefname{listing}{Listing}{Listings}

%\usepackage[newfloat]{minted}

%define IfPackageLoaded
\usepackage{ltxcmds}
\makeatletter
\newcommand{\IfPackageLoaded}[2]{\ltx@ifpackageloaded{#1}{#2}{}}
\makeatother

\IfPackageLoaded{minted}{
  % Line numbers not flowing out of the margin
  \setminted{numbersep=5pt, xleftmargin=12pt}

  %http://www.jevon.org/wiki/Eclipse_Pygments_Style
  %\usemintedstyle{eclipse}
  %
  %\usemintedstyle{autumn}
  %\usemintedstyle{rrt}
  %\usemintedstyle{borland}
  %\usemintedstyle{friendlygrayscale}
  \usemintedstyle{friendly}

  % We need to load caption to have a bold font on the label
  % The other parameters mimic the layout of the LNCS class
  \usepackage[labelfont=bf,font=small,skip=4pt]{caption}
  \SetupFloatingEnvironment{listing}{name=List.,within=none}
}{
}
%
%Following definitions are outside of IfPackageLoaded; inside, they are not visible
%
%Intermediate solution for hyperlined refs. See https://tex.stackexchange.com/q/132420/9075 for more information.
\newcommand{\Vlabel}[1]{\label[line]{#1}\hypertarget{#1}{}}
\newcommand{\lref}[1]{\hyperlink{#1}{\FancyVerbLineautorefname~\ref*{#1}}}

%If minted is not loaded, provide the environments nevertheless
\ifcsmacro{listing}{}{
  \newenvironment{listing}[1][htbp!]{\begin{figure}[#1]}{\end{figure}}
  \newcounter{listing}
}
\ifcsmacro{minted}{}{%
  \newenvironment{minted}[1][]{
    \bfseries
    Package minted was not loaded, so there is no XML code shown.

    In case you load minted, please be sure to \begin{enumerate}[a)]
      \item Have \href{https://www.python.org/downloads/}{python} and \href{http://pygments.org/download/}{pygments} installed
      \item Excecute pdflatex using \texttt{-shell-escape}
    \end{enumerate}
  }{}
}

\usepackage{xspace}
%\newcommand{\eg}{e.\,g.\xspace}
%\newcommand{\ie}{i.\,e.\xspace}
\newcommand{\eg}{e.\,g.,\ }
\newcommand{\ie}{i.\,e.,\ }

%introduce \powerset - hint by http://matheplanet.com/matheplanet/nuke/html/viewtopic.php?topic=136492&post_id=997377
\DeclareFontFamily{U}{MnSymbolC}{}
\DeclareSymbolFont{MnSyC}{U}{MnSymbolC}{m}{n}
\DeclareFontShape{U}{MnSymbolC}{m}{n}{
  <-6>    MnSymbolC5
  <6-7>   MnSymbolC6
  <7-8>   MnSymbolC7
  <8-9>   MnSymbolC8
  <9-10>  MnSymbolC9
  <10-12> MnSymbolC10
  <12->   MnSymbolC12%
}{}
\DeclareMathSymbol{\powerset}{\mathord}{MnSyC}{180}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

%% END COPYING HERE

\usepackage{siunitx}
\usepackage{graphicx}
\usepackage{float}

\begin{document}

\title{Topic categorization in Portuguese news articles}
%If Title is too long, use \titlerunning
%\titlerunning{Short Title}

%Single insitute
\author{André F. Santos}
%If there are too many authors, use \authorrunning
%\authorrunning{First Author et al.}
\institute{\textit{CRACS \& INESC-Porto LA} \\ \textit{Faculty
of Sciences, University of Porto}\\ Porto, Portugal \\
afs@inesctec.pt}

%% Multiple insitutes - ALTERNATIVE to the above
% \author{%
%     Firstname Lastname\inst{1} \and
%     Firstname Lastname\inst{2}
% }
%
%If there are too many authors, use \authorrunning
%  \authorrunning{First Author et al.}
%
%  \institute{
%      Insitute 1\\
%      \email{...}\and
%      Insitute 2\\
%      \email{...}
%}

@@

\begin{keywords}
topic categorization, machine learning, text mining
\end{keywords}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Online news articles first appeared as reprints from traditional
newspapers; nowadays, however, they represent now the primary source
of news for some segments of the population, both in developed and
developing countries (whether consumed directly in the newspaper
website, or indirectly e.g. through a social media application or a
feed catcher)\cite{greer2006evolution,boczkowski2005digitizing,chyi1999access}.

Unofficially known as \emph{the fourth branch of government}, the press
plays a vital role within our society, keeping us informed about
the current state of affairs (at a local and global scale) and acting
as a watchdog for the other three branches (legislative, executive and
judicial). The (lack of) freedom of press and access to the news in a
given country is even often considered an indicator of a lack of
democracy\cite{goode2009social,house2009freedom}.

As such, improving the ways citizens can access the information (view
it, query it and search for it) contained in news articles has the
potential to contribute for a more informed and, ultimately, better,
society\cite{bollinger1988tolerant}.

On the other hand, the last decades have witnessed a fast increase on
the rate of publication of digital text documents. Traditional
document types, such as news articles, scientific papers or books are
now published online along with new formats, such as blog posts or
tweets, each having thousands or millions of new documents published
each day\cite{hilbert2011world,allan2006online}.

Publication is not the only step which has moved to the digital world;
in fact, most often nowadays the whole document lifecycle happens
digitally, with virtual tools available for
researching, writing, styling, publishing and sharing\cite{williams2009personal}.

@@

\begin{figure}[H]
\centerline{\includegraphics[width=.67\columnwidth]{imgs/wireframe}}
\caption{Category classification and suggestion on the client side}
\label{fig:wireframe}
\end{figure}

The simplest way of achieving this categorization is to have the
author of the article manually introducing it (e.g. the journalist
typing it on the news article authoring framework); however, this
solution presents some challenges:
\begin{itemize}
    \item It increases the amount of work the author has to do.
    \item The author might not be sure which categories are available.
    \item The author might not be sure which category is the best
        (e.g. \emph{Economics} vs  \emph{Finance}).
    \item It does not scale -- e.g. if the goal is to categorize an
        existing (large) corpus.
\end{itemize}

@@

\section{Methods}
In order to train and evaluate classification algorithms, we first
needed to choose and obtain a suitable dataset. Preferably, this
dataset should contain documents which were previously categorized,
allowing us to skip the time and effort-consuming task of
manually categorizing the articles ourselves. Once this dataset was
chosen and obtained, we would then clean and prepare it to be used to
train the classifiers.

@@

\begin{figure}[H]
\centerline{\includegraphics[width=.97\columnwidth]{imgs/articles}}
\caption{Total number of articles retrieved for each category}
\label{fig:articles}
\end{figure}

@@

Originally, the dataset was obtained as a large MongoDB collection (more
than 2.5 million entries), containing articles from several Portuguese
and international newspapers. The process needed to transform this
collection into data our classifiers could process required querying
the database, exporting the news articles and splitting them into a
train and an evaluation datasets.

@@

The files were stored in folders, separated by category. For each
category, 700 articles were allocated to the train set, and 200 to the
evaluation set. These preprocessing tasks were accomplished using Bash
and Node.js scripts.

Both datasets were loaded into R using the
\texttt{tm}\footnote{\url{https://cran.r-project.org/web/packages/tm/}}
package. Each was then passed to a function responsible for
preprocessing the text of the articles:

\begin{itemize}
    \item The text was converted to lowercase characters.
    \item The Portuguese stopwords were removed.
    \item Diacritics were converted to their normalized form (e.g.
        \framebox{à} $\rightarrow$ \framebox{a}).
    \item Punctuation signs were removed.
    \item Numbers were removed.
    \item Word were stemmed (e.g. \framebox{conseguiram} $\rightarrow$
        \framebox{consegu}).
    \item White space was removed and the text was tokenized.
\end{itemize}

A document-term matrix was calculated using the
\texttt{DocumentTermMatrix} method from the \texttt{tm} package.  For
each of the 5600 documents present in the train set the matrix included
all the terms which met the following requirements:

\begin{itemize}
    \item The term length was between 3 and 30 characters (to discard
        things like URLs or badly tokenized sentences).
    \item The term appeared in the document at least twice.
    \item The term appeared at least in 10 documents.
\end{itemize}
The algorithm used to weigh the terms in the matrix was
\textit{tf-idf}\cite{christopher2008introduction}. The obtained matrix
presented a sparsity of 99\% and contained 3234 distinct terms.

@@

We used the \texttt{removeSparseTerms} function from the \texttt{tm}
package, and produced three distinct lists of terms:
\begin{description}
    \item[-- $LT_{90}$] contained terms with 90\% or less sparsity.
    \item[-- $LT_{95}$] contained terms whose sparsity was under 95\%.
    \item[-- $LT_{99}$] contained the terms with a sparsity level
        below 99\%.
\end{description}

@@

To create the classifiers, we used 5 well known algorithms: decision
tree (DT), k-nearest neighbors (KNN), naive Bayes (NB), neural network
(NN) and support vector
machine -- with radial kernel (SVM\textsubscript{RK}) and linear
kernel (SVM\textsubscript{LK}).

@@

\begin{table}[H]
    \vspace{-.5cm}
    \caption{Lists size}
\begin{center}
\begin{tabular}{l|SS}
    & \parbox{6em}{\centering Sparsity\\threshold (\%)}  &
    \parbox{5em}{\centering Number of\\terms left}   \\\hline
    $LT_{90}$ &  90                 &  45 \\
    $LT_{95}$ &  95                 & 139 \\
    $LT_{99}$ &  99                 & 912 \\
\end{tabular}
\label{tab:size}
\end{center}
\end{table}

@@

\begin{table}[H]
    \vspace{-.2cm}
    \caption{Execution times for training}
\begin{center}
\begin{tabular}{l|cr}
    & Algorithm  & \parbox{3em}{\centering Training\\time} \\\hline
    $LT_{90}$    & DT         & 12s  \\
    $LT_{95}$    & KNN        &  3m  \\
    $LT_{99}$    & KNN        & 57m  \\
\end{tabular}
\label{tab:exectime}
\end{center}
\end{table}

@@

\begin{table}[H]
    \vspace{-.2cm}
    \caption{Error rates}
\begin{center}
\begin{tabular}{l|rrrrrr}
                 & DT            & KNN           & NB   & NN   & SVM\textsubscript{RK} & SVM\textsubscript{LK} \\ \hline
    $LT_{90}$    & \textbf{0.60} & 0.71          & 0.85 & 0.61 & 0.88                  & 0.88                  \\
    $LT_{95}$    & 0.70          & \textbf{0.56} & 0.83 & 0.58 & 0.87                  & 0.87                  \\
    $LT_{99}$    & 0.70          & \textbf{0.35} & 0.87 & 0.76 & 0.87                  & 0.87                  \\
\end{tabular}
\label{tab:error}
\end{center}
\end{table}

@@

\begin{table}[H]
    \caption{Categories confusion matrix ($LT_{99}$ with KNN)}
\begin{center}
\begin{tabular}{l|rrrrrrrr}
& \rotatebox{45}{science}        & \rotatebox{45}{culture}   &
  \rotatebox{45}{sports}         & \rotatebox{45}{economics} &
  \rotatebox{45}{international}  & \rotatebox{45}{music}     &
  \rotatebox{45}{national}       & \rotatebox{45}{politics} \\\hline
science       & {\bf 127} & 19        & 1          & 6         & 12       & 24        & 11       & 0 \\
culture       & 5         & {\bf 102} & 2          & 2         & 3        & 79        & 6        & 1 \\
sports        & 1         & 3         & {\bf  168} & 7         & 3        & 14        & 4        & 0 \\
economics     & 5         & 3         & 0          & {\bf 146} & 5        & 17        & 14       & 10 \\
international & 12        & 12        & 7          & 17        & {\bf 90} & 35        & 16       & 11 \\
music         & 0         & 2         & 0          & 0         & 2        & {\bf 184} & 1        & 0 \\
national      & 9         & 9         & 9          & 19        & 15       & 31        & {\bf 88} & 21 \\
politics      & 1         & 3         & 0          & 30        & 8        & 17        & 14       & {\bf 127} \\
\end{tabular}
\label{tab:confmat}
\end{center}
\end{table}

@@

The size of the list of terms, for example, gives us an idea of the
memory footprint of a classifier, a parameter which is of the utmost
importance if the goal is to implement a classifier as lightweight as
possible. The training time is also relevant, as shorter training
times give the possibility of retraining the classifiers more often,
allowing them to be updated as the corpus of articles grows in size.

Looking at the actual results obtained and represented in
Tables~\ref{tab:size} and \ref{tab:exectime} we can see that $LT_{90}$
has simultaneously the smallest list of terms (45) and the shortest
training time (12 seconds using the DT algorithm). However, $LT_{90}$
also presents the worst error rates, even when looking at the
algorithm which achieved its best results (0.60 using a DT
classifier).

@@

The results obtained confirm that there is a tradeoff between the size
of the list of terms and the training time, on the one hand, and the
classifier error rate, on the other. However, a list of 912 terms
seems to be an acceptable memory footprint for a client side
classifier; additionally, the higher training time would not present
much problems in this scenario, as the classifier would be trained
beforehand and thus not be visible on the client side.

Given the reasonable values for the training time and the size of the
list of terms, and looking to the error rate obtained in each of tests
performed, we can conclude that the best option was to use the largest
list of terms ($LT_{99}$) and the KNN classifier.

@@

The tasks and results previously described already provide useful
insights into this matter. However, the lowest error rate obtained
(0.35) might still be improved upon, either by leveraging new
algorithms, fine tuning the ones already tested, or by increasing the
train corpus.

We have established that it is in fact possible
to develop a news article classifier which is lightweight enough to be
used (in real time) on the client side of a web application. The
following step will be the actual implementation of the classifier,
probably in the form of a JavaScript library.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{splncs03}
\bibliography{article}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
